Reformulating phylogenetic mixed models to improve flexibility and speed

### 2019-04-08 phyloglmm_ms.tex
fix some typos

\documentclass[12pt]{article}
\usepackage{graphicx}
\usepackage{grffile}
\usepackage{color}
\usepackage{soul}
\usepackage{tabularx}
\usepackage{multirow}
% use \xspace to allow for space after a macro as necessary
\usepackage{xspace}
\usepackage[top=1in, bottom=1in, left=1.25in, right=1.25in]{geometry}
\usepackage{pdfpages}
%% maxwidth is the original width if it is less than linewidth
%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

@@

There are two general challenges in linking the evolutionary process to a phylogenetic comparative statistical framework.
% Phylogenetic signal : the magnitude of phylogenetic correlation
The first challenge is to identify and separate phylogenetic signals from other sources of variation present in the data \citep{blomberg2003testing}.  
% \bmb{prev sentence doesn't seem be quite the right lead-in to the rest of the paragraph? You go on to talk about the tip-variation/residual-variation confounding issue. Maybe replace with something talking more specifically about resolving tip variation?}
The classic phylogenetic regression uses a statistical model that is equivalent to assuming that phylogenetic correlation in the residuals from a regression between two species-level traits arises because the residual variation in the trait that is the dependent variable evolves along the branches of the phylogeny according to a Brownian-motion evolutionary model. 
If the residuals are normally distributed and observed without any additional error or within-species variation, Felsenstein's method of phylogenetically independent contrasts (PICs) \citep{felsenstein1985phylogenies, nicolakakis2000forebrain} is sufficient to account for the phylogenetic correlation.
More recent approaches -- including phylogenetic generalized linear mixed models (PGLMM) \citep{ives2011generalized}, Pagel's $\lambda$ \citep{pagel1999inferring}, and Blomberg's $K$ \citep{blomberg2003testing} --- build upon PICs by considering different (non-Normal) response distributions and by accounting for additional complexities in the evolutionary process. 
In particular, they can partition residual variation into (1) uncorrelated, or independent, residual variation (observation error or tip variation) and (2) phylogenetic signal  (biological/evolutionary process error) \citep{hansen2012interpreting}.
However, if each species' traits are observed more than once, possibly under different conditions, we can potentially distinguish third level of variation; in this case, phylogenetic signal and tip variation are now both part of the evolutionary process error (which we will call intercept-level variation) while the among species residual error is associated with among observations variation within each species.
%\bmb{This is a good point, well expressed. Does PGLMM lump tip variation and within-species variation, or is this primarily an issue with Blomberg/Pagel? Thinking about Boettiger's critique \cite{boettiger2013is} (don't know if it's been peer-reviewed, we could ask him), which might be resolved if we think explicitly about within-species variation \ldots
%\mli{I think it is more like our simulation test. In Boettiger's example, it is treating sister taxa as different species. So it is using a phylogenetic tree of 8 species instead of 4.}
Although many studies include multiple observations per species, phylogenetic analyses rarely take advantage of such observations to partition variablility more finely.
Indeed, many existing methods cannot accommodate multiple observations per species, instead requiring users to collapse the data by taking species averages.
%Despite having multiple observations per tip taxon (species), a typical part of observational or experimental design, but it is rarely seen: existing platforms are unable to fit multiple observations, or unbalanced design data, or both.
% \bmb{OK, but we should be prepared to support this. Could we construct a table of existing approaches/platforms with their capabilities?}

\begin{center}
\begin{figure}[h]
  \includegraphics[scale=0.8]{./phylo_diagram.png}
  \caption{Visual categorization of various forms of variation.}
\label{fig:variation}
\end{figure}
\end{center}

%% FIXME: second sentence
The second challenge is flexibly modeling phylogenetic relations across different levels of complexity.
Classic PCA usually examines the response (trait or distribution) to vary according to the phylogenetic relationship across species, but it is plausible the changes in effect (or levels) of the predictor variables 
may vary according to the phylogenetic relationship across species as well.
%\bmb{maybe we can focus on random slopes here? Need to point out that example studies below are \emph{exceptions} to the general assumption of random intercepts only (how did they manage to do their inference?)}
Several recent studies have incorporated phylogenetic signals beyond species level and looked at species response to phylogenetic signals with changes in environmental factors.
For example, \cite{nowakowski2018phylogenetic} considered phylogenetically correlated slopes in response to habitat conversion when studying the abundance of amphibian species using a Bayesian implementation of phylogenetic GLMM, while \cite{li2017canfun} considered phylogenetically correlated species nested within sites for plant abundance via phylogenetic GLMM approach. 
The tools available for extending phylogenetic relationships to predictor variables (or predictor level variation) are relatively inflexible; thus, computationally sophisticated biologists have turned to more flexible Bayesian approaches, despite their computational burden \citep{hadfield2010mcmc, burkner2016brms}.
Table 1 provides a summary of model complexities, platforms and data constraints for phylogenetic comparative analysis.

@@

%% FIXME: chart looks ugly
\begin{table}[]
\begin{tabular}{|l|l|l|l|}
\hline
\textbf{Model}                                                                                    & \textbf{Method}                                                       & \textbf{Data}                                                                & \textbf{Platform}                                                                              \\ \hline
\multirow{2}{*}{\begin{tabular}[c]{@{}l@{}}Generalized Linear \\ Model (GLM)\end{tabular}}        & Correlated residual                                                   & Single Observation                                                           & \begin{tabular}[c]{@{}l@{}}\pkg{nlme}:gls, \\ \pkg{ape}:pic\end{tabular}                                   \\ \cline{2-4} 
                                                                                                  & \begin{tabular}[c]{@{}l@{}}Residual \\ + phylo intercept\end{tabular} & Single Observation                                                           & \begin{tabular}[c]{@{}l@{}}Pagel's $\lambda$\\ Blomberg's $k$ \\ via \pkg{nlme}:gls\\ \pkg{phylolm}\end{tabular} \\ \hline
\multirow{2}{*}{\begin{tabular}[c]{@{}l@{}}Generalized Linear \\ Mixed Model (GLMM)\end{tabular}} & \multirow{2}{*}{Random effect}                                        & \begin{tabular}[c]{@{}l@{}}Single Observation\\ Balanced Design\end{tabular} & \pkg{pez}                                                                                            \\ \cline{3-4} 
                                                                                                  &                                                                       & Unrestricted                                                                 & \pkg{lme4}, \pkg{glmmTMB}                                                                                  \\ \hline
\multirow{2}{*}{Bayesian GLMM}                                                                    & \multirow{2}{*}{Random effect}                                        & Balanced Design                                                              & \pkg{MCMCglmm}                                                                                       \\ \cline{3-4} 
                                                                                                  &                                                                       & Unrestricted                                                                 & \pkg{brms}                                                                                           \\ \hline
\end{tabular}
\end{table}

In this paper, we will propose an alternative formulation of phylogenetic generalized linear mixed models that is mathematically equivalent to previous approaches, but more flexible.
Our new formulation can be incorporated in any framework that allows for random effect without the need to implement special correlation structures by incorportating phylogenetic structures as part of the mean model. 
%% can be incorporated relatively easily into any mixed-model platform that allows new random-effects design matrices to be specified, and can be easily extended along with the features of mixed model such as multiple observational designs, phylogenetic random-slopes models, and multiple (nested or crossed) random effects.
We will compare our technique coded in R package \pkg{lme4} and \pkg{glmmTMB} with existing R packages (i.e. \pkg{nlme} \citep{pinheiro2014r}, \pkg{phylolm} \citep{ho2014phylolm}, \pkg{pez} \citep{pearse2015pez} \pkg{phyr} (Li et al, unpublished), \pkg{MCMCglmm} \citep{hadfield2010mcmc}, and \pkg{brms} \citep{burkner2016brms}) fitting models to data from simulated model that incorporates phylogenetic signal from both predictors and tips/species, as well as residual variation.
As well as being flexible, we also expect that our approach will be fast, because \pkg{lme4} and \pkg{glmmTMB} use efficient approaches. 
%% FIXME: move to methods/simulations
% In principle, for any given valid method that matches the simulation model should eventually converge to the simulated parameters. 
% We end by discussing opportunities and practicalities of our method and comments on the state of art for this area of research.

@@

\subsection*{Phylogenetic generalized linear mixed model}
Alternatively, one can use the generalized linear mixed effects modelling (GLMM) framework \citep{lynch1991methods}.
The generalized linear mixed effect model is an extension of the classic regression model, a kind of hierarchical linear model that can flexibly incorportate more than one kind of variability through random effects and for non-gaussian responses.
The typical GLMM has the form:
\begin{align}
\bY & \sim \mathbf{f}(\bmu) + \besp \label{eq:glmm1} \\
\bmu & = \bX \bbeta + \bZ \bb  \label{eq:glmm2} \\
\bb & \sim \textrm{MVN}(0, \bSigma(\theta)) \label{eq:glmm3} \\
\besp & \sim \textrm{N}(0,\sigma^{2}_{\epsilon})
\end{align}
where $\bZ$ is an $n \times m$ model matrix for the $n$ -- dimensional vector-valued $m$ predictor variables; $\bb$ (sometimes referred as the "G-side" effect) represents the conditional modes and assumed to be multivariate normally distributed with a variance-covariance matrix given by $\bSigma(\theta)$; $\besp$ (sometimes referred as "R" effect) is an $n \times 1$ vector of independent and normally distributed error terms with variance $\sigma_{\epsilon}^2$.
Analogously, the phylogenetic regression given by (\ref{eq:gls1}) can be represented in the generalized mixed model framework with Gaussian responses by constraining $\bSigma(\theta) = \sigma^2 \bC$ and $\sigma^{2}_{\epsilon} = 0$.

There are serveral advantages in modelling phylogenetic correlated traits in the mixed model framework.
First, random intercepts can allow the response trait to vary across groups/levels within grouping predictors (e.g. patch, site, experiment, and etc). 
Second, it can also allow the response (trait or distribution) to vary according to the phylogenetic relationships across species (i.e. similar species will have similiar responses), and between species (tip variation).
The third type of variation that is often neglected or rarely considered are random slopes.
Random slopes can allow fixed effects (size of the effect in continuous predictor variables) to vary for each species. 
Analogous to modelling the repsonse dependency via phylogenetic relationship among species, phylogenetic random slopes can allow fixed effects to account for the phylogenetic relations (i.e. similar species will have similar size of the effects on the response than distinct pairs).

@@

%% FIXME: need top/side labels
\begin{center}
\begin{figure}[h]
  \includegraphics[scale=0.8,page=1]{./git_push/tree.pdf}
  \caption{Three species phylogenetic tree.}
\label{fig:tree}
\end{figure}
\end{center}

@@

We generated test data based on the random slopes mixed model formulation (\ref{eq:glmm1}, \ref{eq:glmm2}, \ref{eq:glmm3}, \ref{eq:glmm4} ) with a single response variable $\bY$ and a single predictor continuous variable $\bX$ for $n$ = 25, 50, and 100 species.
For simplicity, the response variable $\bY$ is conditional normally distributed (i.e. gaussian distribution with an idenitiy link function) , which is a linear mixed effect model. 
For the first set of simulation, we simulate one observation per species.
% For wider range of comparisons, we simulate one observation ($X_i,Y_i$) per species.
The full simulation model is the following:
\begin{align}
\bY & = (\beta_0 + b_{\mathrm{phy_{int}}}) + (\beta_1 + b_{\mathrm{phy_{slope}}}) X + \besp  \label{eq:ss_sim}\\\
(b_{\mathrm{phy_{int}}}, b_{\mathrm{phy_{slope}}}) & \sim \textrm{MVN} \left( 0, \begin{bmatrix}
\sigma^2_{\mathrm{phy_{int}}} & \sigma_{\mathrm{phy_{int},phy_{slope}}} \\ 
\sigma_{\mathrm{phy_{int},phy_{slope}}} & \sigma^2_{\mathrm{phy_{slope}}}
\end{bmatrix} 
\right) \\ 
\besp & \sim \textrm{N} ( 0 , \sigma_{\epsilon}^2) .
\end{align}
The model contains two fixed effect parameters ($\beta_0$ and $\beta_1$), three random effect parameters (phylogenetic random intercept variance, phylogenetic random slope variance and covariance between phylogenetic random slope and intercept) and residual variance.  
The covariance between phylognetic random intercept and slope measures the correlation of phylogenetic dependency variability of the size in effect and outcome.
Predictor and intercept level random effect of species are not applicable in this simulation setting because there is no variation at the species level (i.e. single observation per species).

@@

Our algorithmic approach is general and could be implemented in a wide range of computational platforms that support independent latent variables as part of a linear or random effects (Z matrix) in generalized linear model. 
We implemented our approach using the R packages \pkg{lme4} \citep{bates2015fitting} and \pkg{glmmTMB} \citep{brooks2017glmmTMB}.
We compare our approach with five other R packages that can fit phylogenetic comparative models: \pkg{nlme} \citep{pinheiro2014r}, \pkg{phylolm} \citep{ho2014phylolm}, \pkg{pez} \citep{pearse2015pez}, and \pkg{brms} \citep{burkner2016brms}.
% \bmb{I think nlme can do other evolutionary models as well? May not be worth mentioning ...}
% \begin{verbatim}
% grep("\\.",apropos("^cor[A-Z]",ignore.case=FALSE),
% invert=TRUE,value=TRUE)
% \end{verbatim}
Phylogenetic generalized least squares (PGLS) (\code{gls} in \pkg{nlme}) is one of the most widely used techniques in phylogenetic comparative analysis; it simply fits a linear model where the covariance structure between species is equivalent to assuming a Brownian-motion evolutionary process (or other evolutionary process) on the tree instead of treating the residual variation as independent and identically distributed. 
Phylogenetic generalized linear models (PGLM) (\code{phyloglm} in \pkg{phylolm}) are a slightly more flexible variation of PGLS that can allow observational residuals and as well as non-Gaussian response variables.
Both packages can model other evolutionary processes and different correlation structures (i.e. Pagel's $\lambda$, Blomberg's $K$ etc.), but we restrict our PGLS fits to the simple BM correlation. 
Neither PGLS nor PGLM can handle multiple observations within a species or random slopes.
One of the few packages that currently fit phylogenetic correlations to predictor level variation is \pkg{pez} (and very recently \pkg{phyr}), which can handle adding additional uncorrelated random slopes and interactions with intercept.
Lastly, Bayesian phylogenetic GLMM using Markov chain Monte Carlo (MCMC) are available and can handle all of the cases described above. 
However, MCMC is often computationally expensive for GLMMs compared to platforms using determinstic optimization.
\pkg{MCMCglmm} \citep{hadfield2010general} is the most commonly used Bayesian phylogenetic GLMM, but we will instead use \pkg{brms}, which uses a more computationally efficient MCMC technique called Hamiltonian Monte Carlo (HMC) \citep{duane1987hybrid} algorithm \citep{li2018fitting}.

% Table \mli{ref something} provides the statistically equivalent fitting capabilities of common/existing phylogenetic methods.
\begin{table}[]
\resizebox{\textwidth}{!}{%
\begin{tabular}{|c|c|c|c|c|c|c|}
\hline
\textbf{Package} & \pkg{nlme} & \pkg{phylolm} & \pkg{lme4}/\pkg{glmmTMB} & \pkg{pez} & \pkg{brms} & \pkg{MCMCglmm} \\ \hline
Single Group & X & X & X &  & X & X \\ \hline
\begin{tabular}[c]{@{}c@{}}Phylo Intercept\\ Phylo Slope\\ Phylo Slope-Intercept correlation\\ Residual\end{tabular} & X & \begin{tabular}[c]{@{}c@{}}X\\ \\ \\ X\end{tabular} & \begin{tabular}[c]{@{}c@{}}X\\ X\\ X\\ X\end{tabular} &  & \begin{tabular}[c]{@{}c@{}}X\\ X\\ X\\ X\end{tabular} & \begin{tabular}[c]{@{}c@{}}X\\ X\\ X\\ X\end{tabular} \\ \hline
Multi-group &  &  & X & X & X & X \\ \hline
\begin{tabular}[c]{@{}c@{}}Phylo Intercept\\ Phylo Slope\\ Phylo Slope-intercept correlation\\ Phylo Species-group interaction\\ Species intercept\\ Species Slope \\ Species Slope-intercept correlation\\ Residual\end{tabular} &  &  & \begin{tabular}[c]{@{}c@{}}X\\ X\\ X\\ X\\ X\\ X\\ X\\ X\end{tabular} & \begin{tabular}[c]{@{}c@{}}X\\ X\\ \\ X\\ X\\ X\\ \\ X\end{tabular} & \begin{tabular}[c]{@{}c@{}}X\\ X\\ X\\ X\\ X\\ X\\ X\\ X\end{tabular} & \begin{tabular}[c]{@{}c@{}}X\\ X\\ X\\ \\ X\\ X\\ X\\ X\end{tabular} \\ \hline
\end{tabular}%
}
\end{table}

%% STOP HERE, MLI will need to do everything below this comment. Everything above this comment is ready for BB to read.

>>>>

\documentclass[12pt]{article}
\usepackage{graphicx}
\usepackage{float}
\usepackage{grffile}
\usepackage{color}
\usepackage{soul}
\usepackage{tabularx}
\usepackage{multirow}
% use \xspace to allow for space after a macro as necessary
\usepackage{xspace}
\usepackage[top=1in, bottom=1in, left=1.25in, right=1.25in]{geometry}
\usepackage{pdfpages}
%% maxwidth is the original width if it is less than linewidth
%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

@@

There are two general challenges in linking the evolutionary process to a phylogenetic comparative statistical framework.
% Phylogenetic signal : the magnitude of phylogenetic correlation
The first challenge is to identify and separate phylogenetic signals from other sources of variation present in the data \citep{blomberg2003testing}.  
% \bmb{prev sentence doesn't seem be quite the right lead-in to the rest of the paragraph? You go on to talk about the tip-variation/residual-variation confounding issue. Maybe replace with something talking more specifically about resolving tip variation?}
The classic phylogenetic regression uses a statistical model that is equivalent to assuming that phylogenetic correlation in the residuals from a regression between two species-level traits arises because the residual variation in the trait that is the dependent variable evolves along the branches of the phylogeny according to a Brownian-motion evolutionary model. 
If the residuals are normally distributed and observed without any additional error or within-species variation, Felsenstein's method of phylogenetically independent contrasts (PICs) \citep{felsenstein1985phylogenies, nicolakakis2000forebrain} is sufficient to account for the phylogenetic correlation.
More recent approaches -- including phylogenetic generalized linear mixed models (PGLMM) \citep{ives2011generalized}, Pagel's $\lambda$ \citep{pagel1999inferring}, and Blomberg's $K$ \citep{blomberg2003testing} --- build upon PICs by considering different (non-Normal) response distributions and by accounting for additional complexities in the evolutionary process. 
In particular, they can partition residual variation into (1) uncorrelated, or independent, residual variation (observation error or tip variation) and (2) phylogenetic signal  (biological/evolutionary process error) \citep{hansen2012interpreting}.
However, if each species' traits are observed more than once, possibly under different conditions, we can potentially distinguish third level of variation; in this case, phylogenetic signal and tip variation are now both part of the evolutionary process error (which we will call intercept-level variation) while the among species residual error is associated with among observations variation within each species.
%\bmb{This is a good point, well expressed. Does PGLMM lump tip variation and within-species variation, or is this primarily an issue with Blomberg/Pagel? Thinking about Boettiger's critique \cite{boettiger2013is} (don't know if it's been peer-reviewed, we could ask him), which might be resolved if we think explicitly about within-species variation \ldots
%\mli{I think it is more like our simulation test. In Boettiger's example, it is treating sister taxa as different species. So it is using a phylogenetic tree of 8 species instead of 4.}
Although many studies include multiple observations per species, phylogenetic analyses rarely take advantage of such observations to partition variability more finely.
Indeed, many existing methods cannot accommodate multiple observations per species, instead requiring users to collapse the data by taking species averages.
%Despite having multiple observations per tip taxon (species), a typical part of observational or experimental design, but it is rarely seen: existing platforms are unable to fit multiple observations, or unbalanced design data, or both.
% \bmb{OK, but we should be prepared to support this. Could we construct a table of existing approaches/platforms with their capabilities?}

\begin{center}
\begin{figure}[H]
  \includegraphics[scale=0.8]{./phylo_diagram.png}
  \caption{Visual categorization of various forms of variation.}
\label{fig:variation}
\end{figure}
\end{center}

%% FIXME: second sentence
The second challenge is flexibly modeling phylogenetic relations across different levels of complexity.
Classic PCA usually examines the response (trait or distribution) to vary according to the phylogenetic relationship across species, but it is plausible the changes in effect (or levels) of the predictor variables 
may vary according to the phylogenetic relationship across species as well.
%\bmb{maybe we can focus on random slopes here? Need to point out that example studies below are \emph{exceptions} to the general assumption of random intercepts only (how did they manage to do their inference?)}
Several recent studies have incorporated phylogenetic signals beyond species level and looked at species response to phylogenetic signals with changes in environmental factors.
For example, \cite{nowakowski2018phylogenetic} considered phylogenetically correlated slopes in response to habitat conversion when studying the abundance of amphibian species using a Bayesian implementation of phylogenetic GLMM, while \cite{li2017canfun} considered phylogenetically correlated species nested within sites for plant abundance via phylogenetic GLMM approach. 
The tools available for extending phylogenetic relationships to predictor variables (or predictor level variation) are relatively inflexible; thus, computationally sophisticated biologists have turned to more flexible Bayesian approaches, despite their computational burden \citep{hadfield2010mcmc, burkner2016brms}.
Table \ref{table:model} provides a summary of model complexities, platforms and data constraints for phylogenetic comparative analysis.

@@

%% FIXME: chart looks ugly
\begin{table}[]
\begin{tabular}{|l|l|l|l|}
\hline
\textbf{Model}                                                                                    & \textbf{Method}                                                       & \textbf{Data}                                                                & \textbf{Platform}                                                                              \\ \hline
\multirow{2}{*}{\begin{tabular}[c]{@{}l@{}}Generalized Linear \\ Model (GLM)\end{tabular}}        & Correlated residual                                                   & Single Observation                                                           & \begin{tabular}[c]{@{}l@{}}\pkg{nlme}:gls, \\ \pkg{ape}:pic\end{tabular}                                   \\ \cline{2-4} 
                                                                                                  & \begin{tabular}[c]{@{}l@{}}Residual \\ + phylo intercept\end{tabular} & Single Observation                                                           & \begin{tabular}[c]{@{}l@{}}Pagel's $\lambda$\\ Blomberg's $k$ \\ via \pkg{nlme}:gls\\ \pkg{phylolm}\end{tabular} \\ \hline
\multirow{2}{*}{\begin{tabular}[c]{@{}l@{}}Generalized Linear \\ Mixed Model (GLMM)\end{tabular}} & \multirow{2}{*}{Random effect}                                        & \begin{tabular}[c]{@{}l@{}}Single Observation\\ Balanced Design\end{tabular} & \pkg{pez}                                                                                            \\ \cline{3-4} 
                                                                                                  &                                                                       & Unrestricted                                                                 & \pkg{lme4}, \pkg{glmmTMB}                                                                                  \\ \hline
\multirow{2}{*}{Bayesian GLMM}                                                                    & \multirow{2}{*}{Random effect}                                        & Balanced Design                                                              & \pkg{MCMCglmm}                                                                                       \\ \cline{3-4} 
                                                                                                  &                                                                       & Unrestricted                                                                 & \pkg{brms}                                                                                           \\ \hline
\end{tabular}
\caption{Model and platforms}
\label{table:model}
\end{table}

In this paper, we will propose an alternative formulation of phylogenetic generalized linear mixed models that is mathematically equivalent to previous approaches, but more flexible.
Our new formulation can be incorporated in any framework that allows for random effect without the need to implement special correlation structures by incorporating phylogenetic structures as part of the mean model. 
%% can be incorporated relatively easily into any mixed-model platform that allows new random-effects design matrices to be specified, and can be easily extended along with the features of mixed model such as multiple observational designs, phylogenetic random-slopes models, and multiple (nested or crossed) random effects.
We will compare our technique coded in R package \pkg{lme4} and \pkg{glmmTMB} with existing R packages (i.e. \pkg{nlme} \citep{pinheiro2014r}, \pkg{phylolm} \citep{ho2014phylolm}, \pkg{pez} \citep{pearse2015pez} \pkg{phyr} (Li et al, unpublished), \pkg{MCMCglmm} \citep{hadfield2010mcmc}, and \pkg{brms} \citep{burkner2016brms}) fitting models to data from simulated model that incorporates phylogenetic signal from both predictors and tips/species, as well as residual variation.
As well as being flexible, we also expect that our approach will be fast, because \pkg{lme4} and \pkg{glmmTMB} use efficient approaches. 
%% FIXME: move to methods/simulations
% In principle, for any given valid method that matches the simulation model should eventually converge to the simulated parameters. 
% We end by discussing opportunities and practicalities of our method and comments on the state of art for this area of research.

@@

\subsection*{Phylogenetic generalized linear mixed model}
Alternatively, one can use the generalized linear mixed effects modelling (GLMM) framework \citep{lynch1991methods}.
The generalized linear mixed effect model is an extension of the classic regression model, a kind of hierarchical linear model that can flexibly incorporate more than one kind of variability through random effects and for non-Gaussian responses.
The typical GLMM has the form:
\begin{align}
\bY & \sim \mathbf{f}(\bmu) + \besp \label{eq:glmm1} \\
\bmu & = \bX \bbeta + \bZ \bb  \label{eq:glmm2} \\
\bb & \sim \textrm{MVN}(0, \bSigma(\theta)) \label{eq:glmm3} \\
\besp & \sim \textrm{N}(0,\sigma^{2}_{\epsilon})
\end{align}
where $\bZ$ is an $n \times m$ model matrix for the $n$ -- dimensional vector-valued $m$ predictor variables; $\bb$ (sometimes referred as the "G-side" effect) represents the conditional modes and assumed to be multivariate normally distributed with a variance-covariance matrix given by $\bSigma(\theta)$; $\besp$ (sometimes referred as "R" effect) is an $n \times 1$ vector of independent and normally distributed error terms with variance $\sigma_{\epsilon}^2$.
Analogously, the phylogenetic regression given by (\ref{eq:gls1}) can be represented in the generalized mixed model framework with Gaussian responses by constraining $\bSigma(\theta) = \sigma^2 \bC$ and $\sigma^{2}_{\epsilon} = 0$.

There are several advantages in modelling phylogenetic correlated traits in the mixed model framework.
First, random intercepts can allow the response trait to vary across groups/levels within grouping predictors (e.g. patch, site, experiment, and etc). 
Second, it can also allow the response (trait or distribution) to vary according to the phylogenetic relationships across species (i.e. similar species will have similar responses), and between species (tip variation).
The third type of variation that is often neglected or rarely considered are random slopes.
Random slopes can allow fixed effects (size of the effect in continuous predictor variables) to vary for each species. 
Analogous to modelling the response dependency via phylogenetic relationship among species, phylogenetic random slopes can allow fixed effects to account for the phylogenetic relations (i.e. similar species will have similar size of the effects on the response than distinct pairs).

@@

%% FIXME: need top/side labels
\begin{center}
\begin{figure}[H]
  \includegraphics[scale=0.8,page=1]{./git_push/tree.pdf}
  \caption{Three species phylogenetic tree.}
\label{fig:tree}
\end{figure}
\end{center}

@@

We generated test data based on the random slopes mixed model formulation (\ref{eq:glmm1}, \ref{eq:glmm2}, \ref{eq:glmm3}, \ref{eq:glmm4} ) with a single response variable $\bY$ and a single predictor continuous variable $\bX$ for $n$ = 25, 50, and 100 species.
For simplicity, the response variable $\bY$ is conditional normally distributed (i.e. Gaussian distribution with an identity link function) , which is a linear mixed effect model. 
For the first set of simulation, we simulate one observation per species.
% For wider range of comparisons, we simulate one observation ($X_i,Y_i$) per species.
The full simulation model is the following:
\begin{align}
\bY & = (\beta_0 + b_{\mathrm{phy_{int}}}) + (\beta_1 + b_{\mathrm{phy_{slope}}}) X + \besp  \label{eq:ss_sim}\\\
(b_{\mathrm{phy_{int}}}, b_{\mathrm{phy_{slope}}}) & \sim \textrm{MVN} \left( 0, \begin{bmatrix}
\sigma^2_{\mathrm{phy_{int}}} & \sigma_{\mathrm{phy_{int},phy_{slope}}} \\ 
\sigma_{\mathrm{phy_{int},phy_{slope}}} & \sigma^2_{\mathrm{phy_{slope}}}
\end{bmatrix} 
\right) \\ 
\besp & \sim \textrm{N} ( 0 , \sigma_{\epsilon}^2) .
\end{align}
The model contains two fixed effect parameters ($\beta_0$ and $\beta_1$), three random effect parameters (phylogenetic random intercept variance, phylogenetic random slope variance and covariance between phylogenetic random slope and intercept) and residual variance.  
The covariance between phylognetic random intercept and slope measures the correlation of phylogenetic dependency variability of the size in effect and outcome.
Predictor and intercept level random effect of species are not applicable in this simulation setting because there is no variation at the species level (i.e. single observation per species).

@@

Our algorithmic approach is general and could be implemented in a wide range of computational platforms that support independent latent variables as part of a linear or random effects (Z matrix) in generalized linear model. 
We implemented our approach using the R packages \pkg{lme4} \citep{bates2015fitting} and \pkg{glmmTMB} \citep{brooks2017glmmTMB}.
We compare our approach with five other R packages that can fit phylogenetic comparative models: \pkg{nlme} \citep{pinheiro2014r}, \pkg{phylolm} \citep{ho2014phylolm}, \pkg{pez} \citep{pearse2015pez}, and \pkg{brms} \citep{burkner2016brms}.
% \bmb{I think nlme can do other evolutionary models as well? May not be worth mentioning ...}
% \begin{verbatim}
% grep("\\.",apropos("^cor[A-Z]",ignore.case=FALSE),
% invert=TRUE,value=TRUE)
% \end{verbatim}
Phylogenetic generalized least squares (PGLS) (\code{gls} in \pkg{nlme}) is one of the most widely used techniques in phylogenetic comparative analysis; it simply fits a linear model where the covariance structure between species is equivalent to assuming a Brownian-motion evolutionary process (or other evolutionary process) on the tree instead of treating the residual variation as independent and identically distributed. 
Phylogenetic generalized linear models (PGLM) (\code{phyloglm} in \pkg{phylolm}) are a slightly more flexible variation of PGLS that can allow observational residuals and as well as non-Gaussian response variables.
Both packages can model other evolutionary processes and different correlation structures (i.e. Pagel's $\lambda$, Blomberg's $K$ etc.), but we restrict our PGLS fits to the simple BM correlation. 
Neither PGLS nor PGLM can handle multiple observations within a species or random slopes.
One of the few packages that currently fit phylogenetic correlations to predictor level variation is \pkg{pez} (and very recently \pkg{phyr}), which can handle adding additional uncorrelated random slopes and interactions with intercept.
Lastly, Bayesian phylogenetic GLMM using Markov chain Monte Carlo (MCMC) are available and can handle all of the cases described above. 
However, MCMC is often computationally expensive for GLMMs compared to platforms using deterministic optimization.
\pkg{MCMCglmm} \citep{hadfield2010general} is the most commonly used Bayesian phylogenetic GLMM, but we will instead use \pkg{brms}, which uses a more computationally efficient MCMC technique called Hamiltonian Monte Carlo (HMC) \citep{duane1987hybrid} algorithm \citep{li2018fitting}.

% Table \mli{ref something} provides the statistically equivalent fitting capabilities of common/existing phylogenetic methods.
\begin{table}[]
\resizebox{\textwidth}{!}{%
\begin{tabular}{|c|c|c|c|c|c|c|}
\hline
\textbf{Package} & \pkg{nlme} & \pkg{phylolm} & \pkg{lme4}/\pkg{glmmTMB} & \pkg{pez} & \pkg{brms} & \pkg{MCMCglmm} \\ \hline
Single Group & X & X & X &  & X & X \\ \hline
\begin{tabular}[c]{@{}c@{}}Phylo Intercept\\ Phylo Slope\\ Phylo Slope-Intercept correlation\\ Residual\end{tabular} & X & \begin{tabular}[c]{@{}c@{}}X\\ \\ \\ X\end{tabular} & \begin{tabular}[c]{@{}c@{}}X\\ X\\ X\\ X\end{tabular} &  & \begin{tabular}[c]{@{}c@{}}X\\ X\\ X\\ X\end{tabular} & \begin{tabular}[c]{@{}c@{}}X\\ X\\ X\\ X\end{tabular} \\ \hline
Multi-group &  &  & X & X & X & X \\ \hline
\begin{tabular}[c]{@{}c@{}}Phylo Intercept\\ Phylo Slope\\ Phylo Slope-intercept correlation\\ Phylo Species-group interaction\\ Species intercept\\ Species Slope \\ Species Slope-intercept correlation\\ Residual\end{tabular} &  &  & \begin{tabular}[c]{@{}c@{}}X\\ X\\ X\\ X\\ X\\ X\\ X\\ X\end{tabular} & \begin{tabular}[c]{@{}c@{}}X\\ X\\ \\ X\\ X\\ X\\ \\ X\end{tabular} & \begin{tabular}[c]{@{}c@{}}X\\ X\\ X\\ X\\ X\\ X\\ X\\ X\end{tabular} & \begin{tabular}[c]{@{}c@{}}X\\ X\\ X\\ \\ X\\ X\\ X\\ X\end{tabular} \\ \hline
\end{tabular}%
}
\caption{Parameters}
\label{table:platform}
\end{table}

\subsection*{Simulation and evaluations}
We simulated 100 phylogenetic trees for each sample size (n = 25,50,100, 500) and incorporated simulation parameters (illustrated in figure 2 and 5) to simulate responses from the simulation model (\ref{eq:ss_sim}, \ref{eq:ms_sim}).
Each realization were fitted to all model variant using different platforms.
Tables \ref{table:platform} shows the parameters that are estimable for each platform. 
We only evaluated the goodness of fit for model fits that passed the convergence tests implemented by the package.
For Bayesian GLMM, we combined two criteria to assess convergence: we required a value of the Gelman and Rubin statistic less than 1.1 and an effective sample size (ESS) greater than 1000 for the fixed effect parameters ($\beta_{0} \ \textrm{and} \ \beta_{1}$) for each replication. 
For each replication we sample two chains starting with 10000 iterations. 
%% TODO MAKE SURE THIS IS CORRECT!
We first evaluate our estimates by looking at the distribution of the estimated values (maximum likelihood estimates for non-Bayesian platforms and posterior medians for Bayesian platforms -- this gives us an idea of bias and variation (i.e. quality of the point estimate).
Then we look at frequentist coverage, to assess quality of confidence intervals.
Coverage refers to the frequency with which the computed confidence intervals include the true values of parameters.
We used 90\% (Wald and profile) confidence intervals and quantile-based intervals (for Bayesian GLMM) to evaluate coverage.
We also compare computational speed between different platforms to evaluate the efficiency of the platforms and methods.
